Physcraper: A Python package for continually updated phylogenetic trees using the Open Tree of Life

### 2021-05-28 sources/bmc_template_submission/bmc_article.tex
some moretiny edits

\section*{Discussion}
Databases preserving and democratizing access to biological data
have become essential resources for science.
New molecular data keep accumulating and tools facilitating its integration into
existent evolutionary knowledge contribute to the acceleration of scientific discovery.

Physcraper is a tool that builds upon previous knowledge stored in published
alignments and phylogenies, taking advantage of OpenTree's services to facilitate comparison
of phylogenies, with the main goal of extending our knowledge of phylogenetic relationships
across the tree of life.

We believe this is a key step to successfully establish an open, reproducible workflow for phylogenetics.
It contributes to the democratization of the field by facilitating access to phylogenetic knowledge for ecologists and other non-specialists.

As a tool for automatizing phylogenetic reconstruction from molecular databases,
Physcraper presents several advantages over existing phylogenetic pipelines designed to
make evolutionary sense of the vast amount of
public molecular data available.

Several analysis tools create full phylogenies \textit{de novo} by mining of molecular
databases \cite{antonelli2017toward, sanderson2008phylota, smith2019pyphlawd,
bennett2018phylotar, pearse2013phylogenerator}.
In particular, Phylota \cite{sanderson2008phylota},
and PHLAWD \cite{smith2009mega}, have been cited and used abundantly.

Physcraper builds on this automated database mining concept by incorporating prior
phylogenetic work and existing taxonomic domain knowledge on appropriate markers
and alignment construction.
This decreases error (requiring less manual downstream processing) and
eases comparison with previous phylogenetic knowledge.

Results from the verification test highlight the importance of incorporating existing expertly
curated homology statements to automatically update phylogenetic relationships, instead
of ignoring the information they contain and building homology statements fully \textit{de novo}.

We encourage users to look at the output files containing information about the
filtered sequences, and potentially modify configuration parameters such as the
sequence length cutoff parameter, based on the filtered sequences.
Default filtering parameters are arbitrary, but we hope that by making the process
of locating homologous sequences online reproducible, and tracking what filters
are used, we make it easier for researchers to delve into the effect that different
choices have on their inferences.
This is in contrast to ``manual" searches for taxa, where similarly arbitrary filters
are applied, but are difficult to trace.
As many studies have shown \cite{huang_what_2009} the effect of missing data can
be enigmatic, and interact with the true phylogenetic relationships for the dataset at hand.
There is not currently strong support in the literature for any particular cutoff
value, and rather than prescribe specific approaches, we encourage users to explore
the effects of different choices on their phylogenetic inferences.
In addition, by providing the output files at each step of the analysis,
it is straightforward to assess how changing parameter and software choices do or
do not drive differences in phylogenetic inference.
By gathering the DNA sequences, and making the unaligned files easily accessible and reusable,
Physcraper facilitates the exploration of alternative aligning tools to determine their effect on phylogenetic inferences.
Once sequences are aligned, users can apply practically any phylogenetic
software and compare results using Physcraper tools.

Organellar genome sequences, such as chloroplasts and mitochondria will also generally
be excluded from automatic addition based on default Physcraper length cutoffs.
Multiple sequence alignment of loci of drastically different lengths is unfeasible,
and we have found in testing that it often returns incorrect results, splitting
shorter sequences with many long gaps to align with exact matches across the entire
longer locus.
While it would be possible to directly extract the BLAST match from genomes, this
 would exclude potentially homologous flanking regions which are not matched
using BLAST's local search algorithm, but that may be important for phylogenetic inference.
Instead we list the accession numbers for these matches in the ``seqlen\_mismatch.txt"
file, for users to assess and incorporate appropriate homologous regions to their
alignment of interest.

Unlike phylogenetic placement approaches \cite{berger_performance_2011, matsen_pplacer_2010},
which add new taxa without modifying the input phylogeny, Physcraper estimates all
the relationships anew in the context of the new data.
PUMPER \cite{izquierdo2014pumper} shares these conceptual strengths, but is no
longer under active development, is challenging to install and run, and has produced
 very few published phylogenetic analyses since its publication.

@@

The Physcraper workflow can be used to rapidly (in a matter of hours) create
phylogenies to address challenges overarching both fields of ecology and evolution, such as
phylogenetically placing newly discovered species \cite{webb2010biodiversity},
curating taxonomic assignments
\cite{san2010molecular},
and generating custom trees for ecological \cite{helmus2012phylogenetic} and
evolutionary downstream analyses \cite{stoltzfus2013phylotastic}.

>>>>

\section*{Discussion}
Databases preserving and democratizing access to biological data
have become essential resources for science. New molecular data keep accumulating and tools facilitating its integration into
existent evolutionary knowledge contribute to the acceleration of scientific discovery.
Physcraper is a tool that builds upon previous knowledge stored in published
alignments and phylogenies, taking advantage of OpenTree's services to facilitate comparison
of phylogenies, with the main goal of extending our knowledge of phylogenetic relationships
across the tree of life. We believe this is a key step to successfully establish an open, reproducible workflow for phylogenetics.
As such, it facilitates access to phylogenetic knowledge for non-specialists in diverse research areas (e.g., ecology, medicine),
effectively contributing to the democratization of phylogenetic studies.

As a tool for automatizing phylogenetic reconstruction from molecular databases,
Physcraper presents several advantages over existing phylogenetic pipelines designed to
make evolutionary sense of the vast amount of
public molecular data available. Unlike phylogenetic placement approaches \cite{berger_performance_2011, matsen_pplacer_2010},
which add new taxa without modifying the input phylogeny, Physcraper estimates all
the relationships anew in the context of the new data.
The tool PUMPER \cite{izquierdo2014pumper} shares these conceptual strengths, but is no
longer under active development, is challenging to install and run, and has produced
few published phylogenetic analyses since its publication.
Several other existing tools create full phylogenies \textit{de novo} by mining of molecular
databases \cite{antonelli2017toward, sanderson2008phylota, smith2019pyphlawd,
bennett2018phylotar, pearse2013phylogenerator}. In particular, Phylota \cite{sanderson2008phylota},
and PHLAWD \cite{smith2009mega}, have been cited and used abundantly.
Physcraper adds to this automated database mining concept by incorporating prior
phylogenetic work and existing taxonomic domain knowledge on appropriate markers
and alignment construction, to update existing phylogenetic knowledge.
This decreases error (requiring less manual downstream processing) and
eases comparison with previous phylogenetic hypotheses.
Results from the verification test highlight the importance of incorporating existing expertly
curated homology statements to automatically update phylogenetic relationships, instead
of ignoring the information they contain and building homology statements fully \textit{de novo}.

We encourage users to look at the output files containing information about the
filtered sequences, and potentially modify configuration parameters such as the
sequence length cutoff parameter, based on said information.
Physcraper's default filtering parameters are arbitrary, but we hope that by making the process
of locating homologous sequences online reproducible, and tracking what filters
are used, we make it easier for researchers to delve into the effect that different
filtering choices have on their inferences.
This is in contrast to ``manual" searches for taxa, where similarly arbitrary filters
are applied, but are difficult to trace.
It has been shown \cite{huang_what_2009} that the effect of missing data in alignments can
be enigmatic, and interact with the true phylogenetic relationships for the dataset at hand.
There is not currently strong support in the literature for any particular filtering cutoff
value, and rather than prescribe specific approaches, we encourage users to explore
the effects of different parameter values on resulting phylogenetic hypotheses.
In addition, by providing the output files at each step of the analysis,
it is straightforward to assess how changing filtering cutoffs and software choices might drive differences in phylogenetic inference.
By gathering the DNA sequences, and making the unaligned files easily accessible and reusable,
Physcraper facilitates the exploration of alternative aligning tools to determine their effect on phylogenetic inferences.
Once sequences are aligned, users can apply practically any phylogenetic
software and compare results using Physcraper tools in a reproducible framework.

Organellar genome sequences, such as chloroplasts and mitochondria will also generally
be excluded from automatic addition based on default Physcraper length cutoffs.
Multiple sequence alignment of loci of drastically different lengths is unfeasible,
and we have found in testing that it often returns incorrect results, splitting
shorter sequences with many long gaps to align with exact matches across the entire
longer locus.
While it would be possible to directly extract the BLAST match from genomes, this
 would exclude potentially homologous flanking regions which are not matched
by BLAST's local search algorithm, but that may be important for phylogenetic inference.
Instead we list the accession numbers for these matches in the ``seqlen\_mismatch.txt"
file, for users to assess and incorporate appropriate homologous regions to their
alignment of interest.

@@

The Physcraper workflow can be used to rapidly (in a matter of hours) create
phylogenies which can address challenges overarching both fields of ecology and evolution, such as
phylogenetically placing newly discovered species \cite{webb2010biodiversity},
curating taxonomic assignments
\cite{san2010molecular},
and generating custom trees for ecological \cite{helmus2012phylogenetic} and
evolutionary downstream analyses \cite{stoltzfus2013phylotastic}.
